A Short Guide To Modular Functors, Anyons, etc.

### 2016-09-22 guide.tex
commit edits

In the physics literature, modular functors are
explicitly used in Refs.~\cite{Freedman2002, Freedman2002simulation, Beverland2014, Kitaev2006topo} (in some of these references modular functors are simply called TQFTs).
%Also, \cite{Beverland2014} and \cite{Kitaev2006topo}
%use the language of modular functors but they call them TQFT's.
%This is in fact reasonable because a modular functor can be
%seen as part of a TQFT, but is quite confusing to the
%novice who attempts to delve into the mathematical literature.

The definition of a modular functor appears to
be well motivated physically.
Unfortunately, there are many such definitions
in the mathematical literature 
\cite{Walker1991,Turaev1994,Bakalov2001,Tillmann1998}.
According to \cite{Bartlett2015} section 1.2 and 1.3,
there are several open questions involved in 
rigorously establishing the connection between these different axiomatizations.
%In particular, what physicists call anyon theory,
%and mathematicians call a modular tensor category,
Furthermore, the concept of modular tensor category 
has not been established to correspond exactly
(bijectively) to any of these modular functor variants. 
We try not to concern ourselves too much with these details, 
but merely note these facts as a warning to the reader
who may go searching for the ``one true formulation'' of
topological quantum field theory.

@@

In this section we will sketch how a 
braided tensor category arises from a modular functor.
In the mathematical literature,
this is covered in \cite{Turaev1994,Tillmann1998,Bakalov2001}
but as we just noted they use a different formulation for a
modular functor.
%The much more ambitious task of deriving a modular
%functor from a braided tensor category is

@@

In three spatial dimensions, the process of winding one
particle around another, a \emph{monodromy}, is topologically trivial.
This is because the path can be deformed back to the
identity; there is no obstruction:
\begin{center}
\includegraphics[]{pic-monodromy3d.pdf}
\end{center}
The square root of this operation is a swap:
\begin{center}
\includegraphics[]{pic-swap.pdf}
\end{center}
For identical particles this is a symmetry of the system. \cggb{What does symmetry mean here? Since a swap of non-bosons will pick up a phase.}
Continuing with this line of thought 
leads to consideration of the \emph{symmetric group}
on $n$ letters, $S_n$. 
This group is generated by the $n-1$ swap operations
$s_1,...,s_{n-1}$ that obey the relations
\begin{align*}
    s_i^2 &= 1, \\
    s_i s_j &= s_j s_i \ \ \ \mbox{for}\ \ |i-j|>1,\\
    s_i s_{i+1} s_i &= s_{i+1} s_i s_{i+1} \ \ \ \mbox{for}\ \ 1\le i \le n-2.
\end{align*}
This group acts on the statespace of $n$ particles $\F_n$
via unitaries $U(\F_n)$:
$$
    \Phi : S_n \to U(\F_n).
$$

@@

What we have given is a topological description of the group $B_n$.
More formally, we can describe these braid world lines as
paths in the \emph{configuration space} of $n$ points.
\newcommand{\Conf}{\mathcal{C}}
\newcommand{\UConf}{\mathcal{UC}}
This space is defined as the product of
a two dimensional space $M$ for each point, minus the subspace where points overlap:
$$
    \Conf_n = \bigl( \prod_{1}^{n} M \bigr) - \Delta, \ \ 
    \Delta = \{(x_1, ..., x_n) | x_i = x_j \ \ \mbox{for some} \ \ i\ne j \}.
$$
Because we are considering identicle particles (so far)
we use the \emph{unlabelled configuration space}
$\UConf_n$ which is the quotient of $\Conf_n$ by the natural action of the
permutation group $S_n:$
\cggb{I was a bit confused by when we care about unlabelled configuration space and when we care about the labelled configuration space. What if we have some but not all particles identical? Do the two formalisms unify by considering some subgroups of the permutation group that preserve particle label or something?}
$$
    \UConf_n = \Conf_n / S_n.
$$
The \emph{geometric braid group} as we have so far informally described it 
can now be rigorously defined as the fundamental group:
$$
    B_n = \pi_1 ( \UConf_n, x )
$$
where $x$ is some reference configuration in $\UConf_n.$
See \cite{Ghrist2014} for further discussion.

@@

There is another important geometric representation of
the braid group which is purely two-dimensional.
%We denote the two-dimensional space 
%The two dimensional space within which the particles reside we denote 
We pick an $n$-point finite subset $Q_n \subset M$
and consider diffeomorphisms \cggb{why diffeomorphisms and not homeomorphisms?} $f : M \to M$
that map $Q_n$ to itself.
The \emph{mapping class group} of $M$ relative to $Q_n$
is the set of all such diffeomorphisms up to an equivalence relation $\sim_{\mbox{iso}}$:
$$
    MCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(Q_n)=Q_n \} / \sim_{\mbox{iso}}
$$
The equivalence relation $\sim_{\mbox{iso}}$ is called \emph{isotopy}
which allows for any continuous deformation of $f:M\to M$ that
fixes each point in $Q_n.$
Each generator $\sigma_i$ of the braid group is found in $MCG(M, Q_n)$
as a \emph{half-twist} that swaps two points $i, i+1 \in Q_n$:
\begin{center}
\includegraphics[]{pic-halftwist.pdf}
\end{center}
In order to show
the action of
this half-twist we have 
decorated the manifold with
a a checkered pattern,
but there is a more important
object that lives on the manifold
itself.
An \emph{observable} is a simple
closed curve in $M$ that does
not intersect $Q_n.$
Such a closed curve is called an
observable because these will be
associated to measurements
that measure the charge on the
interior of the curve.
The importance of understanding the
braid group as identicle to the mapping class group
is now manifest:
whereas geometric braids act on states
as in a Schr\"{o}dinger picture,
elements of the mapping class group
act on the observables as in
a Heisenberg picture.
\begin{center}
\includegraphics[]{pic-heisenberg.pdf}
\end{center}
This diagram should give the reader some idea as
to why these two definitions of the braid group are
equivelant, but the actual proof of this is somewhat involved.
We refer the reader to \cite{Kassel2010} for an excellent contemporary
reference that fills in these gaps.

@@

Here we show four 
geometric approaches to representing a twist.
Anyone that has struggled to untangle their headphone cable
will immediately see what is going on here.
On the left we have a loop; it is not a braid because it
travels backwards in time.
If we pull on this loop to make it straight, we introduce a twist.
This is shown in the next figure, 
where we show a \emph{framing} which is a non-degenerate \cggb{What's a non-degenerate vector field? What goes wrong if its degenerate?}
vector field along the world-lines of a braid.
The initial and final vectors in the vector field must be the same.
In the next figure we show a \emph{ribbon}:
instead of point particles we have short one-dimensional curves in $M$.
%This is another way to encode a twist without using the differential
%structure of $M.$
On the right the particle is represented as a boundary
component of $M$ with a distinguished point.
In this picture the world-line looks like a tube.
%It is not difficult to see that these 
All these representations of twists carry
essentially the same information.
\begin{center}
\includegraphics[]{pic-framed.pdf}
\end{center}
In the sequal we will stick
to thinking of particles as boundary
components because this fits well
with the way we are formulating 
observables as simple closed curves in $M$.
%More concretely,
%we augment the configuration space
%with tangent vectors. For each point $x_i$
%in a configuration $(x_1,...,x_n)\in \prod M$ we
%attach a non-zero tangent vector $v_i\in T_{x_i}M.$
%In terms of a mapping class group, we will
%consider not just a manifold $M$ with
%distinguished points $Q_n,$ but a manifold with $n$ disjoint
%boundary componends, or ``holes''.
%Each of these boundary components comes
%with a distinguished point.

@@

By a \emph{map of surfaces} $f:M\to N$ we mean 
a diffeomorphism that preserves
manifold orientation, hole labels, and base points.
%\cggb{If we require that the map preserves base points and so on, are we restricting to maps between manifolds with the same number of holes?}
%\simon{Yes, a diffeomorphism is also a homeomorphism, so these manifolds are
%topologically equivalent.}
Note that we also deal with maps of various other objects
(vector spaces, sets, etc.)
but a map of surfaces will have these specific
requirements.

%By an \emph{isotopy} of $f:M\to N$ we mean a 
%map $g:M\to N$
%that is a continuos deformation of $f.$
Two maps of surfaces $f:M\to N$ and $f':M\to N$
will be called \emph{isotopic}
when one is a continuous deformation of the other.
In detail,
we have a continuously
parametrized family of maps $f_t:M\to N$ for $t\in [0, 1]$
such that $f_0=f,\ f_1=f'$
and the restriction of $f_t$ to the 
set of marked points $X\subset\partial M$ is constant:
$f_t|_{X} = f|_{X}$ for $t\in [0, 1].$
%and the restriction of $f_t$ to the boundary is constant:
%$f_t|_{\partial M} = f|_{\partial M}$ for $t\in [0, 1].$
Such a family $\{f_t\}_{t\in [0,1]}$ is called an \emph{isotopy}
of $f.$
This is an equivalence relation on maps $M\to N$, and the equivalence
class of $f$ under isotopy is called the \emph{isotopy class} of $f.$ \cggb{I didn't quite see why this more formal definition is the same as equivalent up to continuous deformation. Why can I restrict to maps where the base points are fixed along the whole path?}

@@

{\bf Unit axioms.}
The fusion space of an empty surface is one dimensional, 
$\F(\phi) \cong \C.$ % preserves the identity of tensor
%\cggb{Does a modular functor have to be over $\C$?}
%\simon{No, mathematicians do this over any ring $k.$}
For $M_a$ a disc with boundary label $a$ we have
$\F(M_\vac)\cong \C$ and 
$\F(M_a)\cong 0$ for $a\ne \vac.$
For an annulus $M_{a,b}$ with boundary labels
$a, b$ we have
$\F(M_{a,\widehat{a}})\cong \C$ and 
$\F(M_{a,b})\cong 0$ for $a\ne \widehat{b}.$

@@

\subsub{POP decomposition.}
%The reverse process of cutting a surface up
%into pair-of-pants induces a direct sum decomposition
%of the fusion space.
Given a manifold $M$ which is
a disc with two or more holes, 
we show how to present $M$ as the gluing of various
pair-of-pants. Such an arangement will be termed a \emph{POP decomposition.}
(We refer to \cite{Ivanov2001} for more details on this construction,
and \cite{Ghrist2014} for a leisurely description of Morse theory.)
This will yield a decomposition of $\F(M)$ into a direct
sum of fusion spaces of pair-of-pants. % $V_{ab}^c$'s.
%The pair-of-pants can be found using Morse theory.
%We proceed using Morse theory.
%Note that we need $M$ to be a differentiable manifold for
%this to work.
The key idea is to choose a ``height'' function $h:M\to\R$
with some specific properties that allow us to
cut the manifold up along level sets of $h.$
First, we need that critical points of $h$ are isolated.
This is the defining condition for $h$ to be a \emph{Morse function}.
%Choose a Morse function (a generic ``height'' function) $h:M\to\R$
Also, we need that $h$ is constant on $\partial M,$
and the values of $h$ at different critical points are distinct.
Now choose a sequence of non-critical values $a_1<a_2<...<a_n$ in $\R$
such that every interval $[a_{i-1}, a_i]$ 
contains exactly one critical value of $h$ and the image
of $h$ lies within $[a_1, a_n].$
%cuts the manifold into pieces $f^{-1}[a,b]$.
Each component of $h^{-1}([a_{i-1}, a_i])$ is then either
an annulus, a disc, or a pair-of-pants depending on the index of (any)
critical point it contains.
We then re-glue any annuli or discs until there are no more of these
and we have only pair-of-pants. \cggb{Is this always possible? What if I have something like a cylinder? Then I just get a bunch of annuli however I cut.}
% note: we may still get annuli in a *component* of the pre-image.
\begin{center}
\includegraphics[]{pic-pants.pdf}\ \ \ \ \ \ \ \ \includegraphics[]{pic-pants-1.pdf}
\end{center}

@@

One way to achieve this is via Cerf theory, which is the theory
of how one may deform Morse functions into other Morse functions
and the kind of transitions involved in their critical point structure.
This was the approach used in \cite{Freedman2002simulation}.
In this work we use a simpler method,
which is essentially the same as {skein theory.} \cggb{Is all this curve diagrams and refactoring theorem and and so on original work, or is it collecting known facts? If its original, this should be an independent paper because otherwise nobody from the math phys community will ever find it. Actually maybe in either case it should be an independent arxiv post at least so that more people find and learn from it.}
%Read on to find out how this works!
This is the {refactoring theorem} that we describe below.

@@

\subsub{Standard surfaces.}
For each $n=0,1,...$ and every ordered sequence of
anyon labels $a_1,...,a_n,b$ we choose a \emph{standard surface}.
This is a surface with
$n+1$ boundary components labeled $a_1,...,a_n,{b}$ 
which we denote $\StdM.$ 
For concreteness, we 
let this surface be the subset of
the complex plane defined by
$\bigl\{x\in\C \ \mathrm{st.}\  |x-\frac{1}{2}|\le \frac{1}{2}\bigr\}$
with $n$ discs
$\bigl\{x\in\C \ \mathrm{st.}\  |x-\frac{2i-1}{2n}|< \frac{1}{4n}\bigr\}_{i=1,...n}$
removed. \cggb{Why is this easier to describe on $\mathbb{C}$ rather than $\mathbb{R}^2$? If its all the same, I think it could be safer to rephrase on $\R^2$.}
We then label the interior holes $a_1,...,a_n$ in order of
increasing real coordinate, and the exterior hole is labeled ${b}$.
The base points are placed in the direction of the negative imaginary axis.
As a notational convenience we highlight the \emph{equator}
of the surface %$\StdM$ 
which is the intersection of
the real axis with the surface:
\begin{center}
\includegraphics[]{pic-disc-standard.pdf}
\end{center}

@@

%We next define the gluing of two standard surfaces
We next show how to glue the exterior hole of
a standard surface $M$
to an interior hole of another standard surface $N.$
To do this within $\C$ we rescale and translate the two surfaces so
that the exterior hole of $M$ coincides with the interior hole of $N.$
At this point the union is not in general going to
produce another standard surface,
%However, we can ... \simon{fix this}
and so we remedy this by applying any isotopy within $\C$
that fixes the real axis while moving the surface to a standard surface. \cggb{Does "fix the real" axis here mean that it fixes every point on the real axis, or just maps the real axis to itself?}
%while moving the holes so as to produce
%a standard surface.

\subsub{Curve diagram.}
We next study maps from standard surfaces to arbitrary surfaces. \cggb{I was a bit confused what this means. Maps means between surfaces with the same holes and base points, right? So then don't we only care about maps from standard surfaces to standard surfaces? Is it maybe more like maps from equatorial curve diagrams on standard surfaces to arbitrary curve diagrams on standard surfaces that we want?}
The set of all maps from $\StdM$ 
to a surface $N$ will be denoted as $\Hom(\StdM, N).$ 
We call each such map a \emph{curve diagram},
or more specifically, a curve diagram on $N$.
The reason for this terminology is that we can reconstruct (up to isotopy)
any map $f:\StdM\to N$ from
the restriction of $f$ to the equator of $\StdM.$ \cggb{citation? or obvious?}
In other words, we can uniquely specify any map
$f:\StdM\to N$ by indicating the action of
$f$ on the equator. We take full advantage of this
fact in our notation: any figure of a surface
with a ``green line'' drawn therein is actually notating a 
diffeomorphism, not a surface!

Any curve diagram will act on a standard POP decomposition
of $\StdM$ sending it to a POP decomposition of $N.$
Surprisingly, the converse of this statement also holds:
any POP decomposition of $N$ (a disc with holes)
comes from some curve diagram
acting on a standard POP decomposition.
%\cggb{This is true for any $N$ that is homeomorphic to a punctured disc?}
%\simon{Right, I should be more clear about when exactly we specialize to discs.}
The proof of this is constructive, and we call this
the refactoring theorem below. \cggb{citation? or original work?} %\simon{XXX} 
%The proof of this is constructive and forms the heart of
%our algorithm below.

%\simon{show how to glue two curve diagrams together to get another curve diagram}
The operation of 
gluing of surfaces can be extended to gluing of curve diagrams as
long as we are careful with the way we identify along the seam:
the identification map needs to respect the equator of the curve
diagram. \cggb{Is respecting the equator the same as respecting whatever canonical curve diagram we choose?}

@@

We note in passing two connections to the mathematical literature.
Such curve diagrams have been 
used in the study of braid groups \cite{Dehornoy2002}, and this
is where the name comes from, although our
curve diagrams respect the base points and so could be further
qualified as ``framed'' curve diagrams.
And, we note the similarity of curve diagrams and associated
modular functor 
to the definition of a planar algebra \cite{Jones1999},
the main difference being that planar algebras allow for 
not just two but any even number of curve intersections at each hole. \cggb{What's a planar algebra and is there an easier place for me to learn about it than from the Jones paper?}

@@

%\subsub{Refactoring theorem.}
\subsection{Refactoring theorem}
In this section we specialize to considering planar surfaces only.
The theorem we are building towards shows that
the $R$-moves act transitively \cggb{what do we mean here by transitively} on curve diagrams.
%\simon{We don't much care about transitivity. Should specialize this.}
That is, given two curve diagrams $f$ and $f'$ on $N$ we can find
a sequence of $R$-moves that transform $f$ to $f'.$
%We consider surface $N$ with $n$ boundary components
%$a_1,...,a_n.$
The key idea is to consider the (directed) image of the equator under
%the curve diagrams $f$ and $f'$.
these curve diagrams.
As mentioned previously, this image is sufficient to
define the entire diffeomorphism (up to isotopy) and so we are free
to work with just this image, or \emph{curve},
and we confuse the distinction between 
a curve diagram (a diffeomorphism)
and its curve.
%which we refer to as a \emph{curve.}
%Each such image will touch all the holes of $N.$

>>>>

In the physics literature, modular functors are
explicitly used in Refs.~\cite{Freedman2002, Freedman2002simulation}.
Also, \cite{Beverland2014} and \cite{Kitaev2006topo}
use the language of modular functors but they call them TQFT's.
This is in fact reasonable because a modular functor can be
seen as part of a TQFT, but is quite confusing to the
novice who attempts to delve into the mathematical literature.

The definition of a modular functor appears to
be well motivated physically.
Unfortunately, there are many such definitions
in the mathematical literature 
\cite{Walker1991,Turaev1994,Bakalov2001,Tillmann1998}.
According to \cite{Bartlett2015} section 1.2 and 1.3,
there are several open questions involved in 
rigorously establishing the connection between these different axiomatizations.
In particular, what physicists call anyon theory,
and mathematicians call a modular tensor category,
has not been established to correspond exactly
(bijectively) to any of these modular functor variants. 
We try not to concern ourselves too much with these details, 
but merely note these facts as a warning to the reader
who may go searching for the ``one true formulation'' of
topological quantum field theory.

@@

The main goal of the present work is to sketch how a
braided tensor category arises from a modular functor.
In the mathematical literature,
this is covered in \cite{Turaev1994,Tillmann1998,Bakalov2001}
but as we just noted they use a different formulation for a
modular functor.
%The much more ambitious task of deriving a modular
%functor from a braided tensor category is

@@

In three spatial dimensions, the process of winding one
particle around another, a \emph{monodromy}, is topologically trivial.
This is because the path can be deformed back to the
identity; there is no obstruction:
\begin{center}
\includegraphics[]{pic-monodromy3d.pdf}
\end{center}
The square root of this operation is a swap:
\begin{center}
\includegraphics[]{pic-swap.pdf}
\end{center}
For identical particles this is a symmetry of the system. 
%%%%%% \cggb{What does symmetry mean here? Since a swap of non-bosons will pick up a phase.}
Continuing with this line of thought 
leads to consideration of the \emph{symmetric group}
on $n$ letters, $S_n$. 
This group is generated by the $n-1$ swap operations
$s_1,...,s_{n-1}$ that obey the relations
\begin{align*}
    s_i^2 &= 1, \\
    s_i s_j &= s_j s_i \ \ \ \mbox{for}\ \ |i-j|>1,\\
    s_i s_{i+1} s_i &= s_{i+1} s_i s_{i+1} \ \ \ \mbox{for}\ \ 1\le i \le n-2.
\end{align*}
This group acts on the statespace of $n$ particles $\F_n$
via unitaries $U(\F_n)$:
$$
    \Phi : S_n \to U(\F_n).
$$

@@

What we have given is a topological description of the group $B_n$.
More formally, we can describe these braid world lines as
paths in the \emph{configuration space} of $n$ points.
\newcommand{\Conf}{\mathcal{C}}
\newcommand{\UConf}{\mathcal{UC}}
This space is defined as the product of
a two dimensional space $M$ for each point, minus the subspace where points overlap:
$$
    \Conf_n = \bigl( \prod_{1}^{n} M \bigr) - \Delta, \ \ 
    \Delta = \{(x_1, ..., x_n) | x_i = x_j \ \ \mbox{for some} \ \ i\ne j \}.
$$
Because we are considering identicle particles (so far)
we use the \emph{unlabelled configuration space}
$\UConf_n$ which is the quotient of $\Conf_n$ by the natural action of the
permutation group $S_n:$
%\cggb{I was a bit confused by when we care about
%unlabelled configuration space and when we care about the labelled
%configuration space. What if we have some but not all
%particles identical? Do the two formalisms unify by considering some
%subgroups of the permutation group that preserve particle label or something?}
$$
    \UConf_n = \Conf_n / S_n.
$$
The \emph{geometric braid group} as we have so far informally described it 
can now be rigorously defined as the fundamental group:
$$
    B_n = \pi_1 ( \UConf_n, x )
$$
where $x$ is some reference configuration in $\UConf_n.$
See \cite{Ghrist2014} for further discussion.

@@

There is another important geometric representation of
the braid group which is purely two-dimensional.
We pick an $n$-point finite subset $Q_n \subset M$
and consider diffeomorphisms 
%\cggb{why diffeomorphisms and not homeomorphisms?}
$f : M \to M$
that map $Q_n$ to itself.
The \emph{mapping class group} of $M$ relative to $Q_n$
is the set of all such diffeomorphisms up to an equivalence relation $\sim_{\mbox{iso}}$:
$$
    MCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(Q_n)=Q_n \} / \sim_{\mbox{iso}}
$$
The equivalence relation $\sim_{\mbox{iso}}$ is called \emph{isotopy}
which allows for any continuous deformation of $f:M\to M$ that
fixes each point in $Q_n.$
Each generator $\sigma_i$ of the braid group is found in $MCG(M, Q_n)$
as a \emph{half-twist} that swaps two points $i, i+1 \in Q_n$:
\begin{center}
\includegraphics[]{pic-halftwist.pdf}
\end{center}
In order to show
the action of
this half-twist we have 
decorated the manifold with
a a checkered pattern,
but there is a more important
object that lives on the manifold
itself.
An \emph{observable} is a simple
closed curve in $M$ that does
not intersect $Q_n.$
Such a closed curve is called an
observable because these will be
associated to measurements
that measure the charge on the
interior of the curve.
The importance of understanding the
braid group as identicle to the mapping class group
is now manifest:
whereas geometric braids act on states
as in a Schr\"{o}dinger picture,
elements of the mapping class group
act on the observables as in
a Heisenberg picture.
\begin{center}
\includegraphics[]{pic-heisenberg.pdf}
\end{center}
This diagram should give the reader some idea as
to why these two definitions of the braid group are
equivelant, but the actual proof of this is somewhat involved.
We refer the reader to \cite{Kassel2010} for an excellent contemporary
reference that fills in these gaps.

@@

Here we show four 
geometric approaches to representing a twist.
Anyone that has struggled to untangle their headphone cable
will immediately see what is going on here.
On the left we have a loop; it is not a braid because it
travels backwards in time.
If we pull on this loop to make it straight, we introduce a twist.
This is shown in the next figure, 
where we show a \emph{framing} which is a non-degenerate 
vector field along the world-lines of a braid.
The initial and final vectors in the vector field must be the same.
By non-degenerate we mean that the vector field is everywhere
non-zero and also not tangent to the world-line.
In the next figure we show a \emph{ribbon}:
instead of point particles we have short one-dimensional curves in $M$.
%This is another way to encode a twist without using the differential
%structure of $M.$
On the right the particle is represented as a boundary
component of $M$ with a distinguished point.
In this picture the world-line looks like a tube.
%It is not difficult to see that these 
All these representations of twists carry
essentially the same information.
\begin{center}
\includegraphics[]{pic-framed.pdf}
\end{center}
In the sequal we will stick
to thinking of particles as boundary
components because this fits well
with the way we are formulating 
observables as simple closed curves in $M$.
%More concretely,
%we augment the configuration space
%with tangent vectors. For each point $x_i$
%in a configuration $(x_1,...,x_n)\in \prod M$ we
%attach a non-zero tangent vector $v_i\in T_{x_i}M.$
%In terms of a mapping class group, we will
%consider not just a manifold $M$ with
%distinguished points $Q_n,$ but a manifold with $n$ disjoint
%boundary componends, or ``holes''.
%Each of these boundary components comes
%with a distinguished point.

@@

By a \emph{map of surfaces} $f:M\to N$ we mean 
a diffeomorphism that preserves
manifold orientation, hole labels, and base points.
Note that we also deal with maps of various other objects
(vector spaces, sets, etc.)
but a map of surfaces will have these specific
requirements.

%By an \emph{isotopy} of $f:M\to N$ we mean a 
%map $g:M\to N$
%that is a continuos deformation of $f.$
Two maps of surfaces $f:M\to N$ and $f':M\to N$
will be called \emph{isotopic}
when one is a continuous deformation of the other.
In detail,
we have a continuously
parametrized family of maps $f_t:M\to N$ for $t\in [0, 1]$
such that $f_0=f,\ f_1=f'$
and the restriction of $f_t$ to the 
set of marked points $X\subset\partial M$ is constant:
$f_t|_{X} = f|_{X}$ for $t\in [0, 1].$
%and the restriction of $f_t$ to the boundary is constant:
%$f_t|_{\partial M} = f|_{\partial M}$ for $t\in [0, 1].$
Such a family $\{f_t\}_{t\in [0,1]}$ is called an \emph{isotopy}
of $f.$
This is an equivalence relation on maps $M\to N$, and the equivalence
class of $f$ under isotopy is called the \emph{isotopy class} of $f.$
%\cggb{I didn't quite see why this more formal definition is
%the same as equivalent up to continuous deformation. Why can
%I restrict to maps where the base points are fixed along the whole path?}

@@

{\bf Unit axioms.}
The fusion space of an empty surface is one dimensional, 
$\F(\phi) \cong \C.$ % preserves the identity of tensor
%\cggb{Does a modular functor have to be over $\C$?}
%\simon{No, mathematicians do this over any field $k.$}
For $M_a$ a disc with boundary label $a$ we have
$\F(M_\vac)\cong \C$ and 
$\F(M_a)\cong 0$ for $a\ne \vac.$
For an annulus $M_{a,b}$ with boundary labels
$a, b$ we have
$\F(M_{a,\widehat{a}})\cong \C$ and 
$\F(M_{a,b})\cong 0$ for $a\ne \widehat{b}.$

@@

\subsub{POP decomposition.}
%The reverse process of cutting a surface up
%into pair-of-pants induces a direct sum decomposition
%of the fusion space.
Given a manifold $M$ which is
a disc with two or more holes, 
we show how to present $M$ as the gluing of various
pair-of-pants. Such an arangement will be termed a \emph{POP decomposition.}
(We refer to \cite{Ivanov2001} for more details on this construction,
and \cite{Ghrist2014} for a leisurely description of Morse theory.)
This will yield a decomposition of $\F(M)$ into a direct
sum of fusion spaces of pair-of-pants. % $V_{ab}^c$'s.
%The pair-of-pants can be found using Morse theory.
%We proceed using Morse theory.
%Note that we need $M$ to be a differentiable manifold for
%this to work.
The key idea is to choose a ``height'' function $h:M\to\R$
with some specific properties that allow us to
cut the manifold up along level sets of $h.$
First, we need that critical points of $h$ are isolated.
This is the defining condition for $h$ to be a \emph{Morse function}.
%Choose a Morse function (a generic ``height'' function) $h:M\to\R$
Also, we need that $h$ is constant on $\partial M,$
and the values of $h$ at different critical points are distinct.
Now choose a sequence of non-critical values $a_1<a_2<...<a_n$ in $\R$
such that every interval $[a_{i-1}, a_i]$ 
contains exactly one critical value of $h$ and the image
of $h$ lies within $[a_1, a_n].$
%cuts the manifold into pieces $f^{-1}[a,b]$.
Each component of $h^{-1}([a_{i-1}, a_i])$ is then either
an annulus, a disc, or a pair-of-pants depending on the index of (any)
critical point it contains.
We then re-glue any annuli or discs until there are no more of these
and we have only pair-of-pants.
%\cggb{Is this always possible? What if I have something like
%a cylinder? Then I just get a bunch of annuli however I cut.}
% note: we may still get annuli in a *component* of the pre-image.
\begin{center}
\includegraphics[]{pic-pants.pdf}\ \ \ \ \ \ \ \ \includegraphics[]{pic-pants-1.pdf}
\end{center}

@@

One way to achieve this is via Cerf theory, which is the theory
of how one may deform Morse functions into other Morse functions
and the kind of transitions involved in their critical point structure.
This was the approach used in \cite{Freedman2002simulation}.
In this work we use a simpler method,
which is essentially the same as {skein theory.}
%    \cggb{Is all this curve diagrams and refactoring theorem and and
%    so on original work, or is it collecting known facts?
%    If its original, this should be an independent paper because
%    otherwise nobody from the math phys community will ever find
%    it. Actually maybe in either case it should be an
%    independent arxiv post at least so that more people find and learn from it.}
%Read on to find out how this works!
This is the {refactoring theorem} that we describe below.

@@

\subsub{Standard surfaces.}
For each $n=0,1,...$ and every ordered sequence of
anyon labels $a_1,...,a_n,b$ we choose a \emph{standard surface}.
This is a surface with
$n+1$ boundary components labeled $a_1,...,a_n,{b}$ 
which we denote $\StdM.$

%For concreteness, we 
%let this surface be the following subset of
%the real plane:
For concreteness
we define this surface using the following 
expression for a closed disc less $n$ open discs:
$$
\Bigl\{(x,y)\in\R^2 \ \mathrm{st.}\  \Bigl|(x,y)-\Bigl(\frac{1}{2},0\Bigr)\Bigr|\le \frac{1}{2}\Bigr\} -
\Bigl\{(x,y)\in\R^2 \ \mathrm{st.}\  \Bigl|(x,y)-\Bigl(\frac{2i-1}{2n},0\Bigr)\Bigr| < \frac{1}{4n}\Bigr\}_{i=1,...n}.
$$
%For concreteness, we 
%let this surface be the subset of
%the complex plane defined by
%$\bigl\{x\in\C \ \mathrm{st.}\  |x-\frac{1}{2}|\le \frac{1}{2}\bigr\}$
%with $n$ discs
%$\bigl\{x\in\C \ \mathrm{st.}\  |x-\frac{2i-1}{2n}|< \frac{1}{4n}\bigr\}_{i=1,...n}$
%removed.
%\cggb{Why is this easier to describe on $\mathbb{C}$ rather than
%$\mathbb{R}^2$? If its all the same, I think it could
%be safer to rephrase on $\R^2$.}
%We then label the interior holes $a_1,...,a_n$ in order of
%increasing real coordinate, and the exterior hole is labeled ${b}$.
%The base points are placed in the direction of the negative imaginary axis.
%As a notational convenience we highlight the \emph{equator}
%of the surface %$\StdM$ 
%which is the intersection of
%the real axis with the surface:
We then label the interior holes $a_1,...,a_n$ in order of
increasing $x$ coordinate, and the exterior hole is labeled $b$.
The base points are placed in the direction of negative $y$ coordinate.
As a notational convenience we highlight the \emph{equator}
of the surface %$\StdM$ 
which is the intersection of
the $x$ axis with the surface:
\begin{center}
\includegraphics[]{pic-disc-standard.pdf}
\end{center}

@@

%We next define the gluing of two standard surfaces
We next show how to glue the exterior hole of
a standard surface $M$
to an interior hole of another standard surface $N.$
To do this within $\R^2$ we rescale and translate the two surfaces so
that the exterior hole of $M$ coincides with the interior hole of $N.$
At this point the union is not in general going to
produce another standard surface,
%However, we can ... \simon{fix this}
and so we remedy this by applying any isotopy within $\R^2$
that fixes the $x$-axis while moving the surface to a standard surface.
%\cggb{Does "fix the real" axis here mean that it fixes
%every point on the real axis, or just maps the
%real axis to itself?}

\subsub{Curve diagram.}
We next study maps from standard surfaces to arbitrary surfaces.
%\cggb{I was a bit confused what this means. Maps means
%between surfaces with the same holes and base points, right?
%So then don't we only care about maps from standard
%surfaces to standard surfaces? Is it maybe more like maps
%from equatorial curve diagrams on standard surfaces to arbitrary curve
%diagrams on standard surfaces that we want?}
The reader should think of this as akin to choosing a basis
for a vector space.
The set of all maps from $\StdM$ 
to a surface $N$ will be denoted as $\Hom(\StdM, N).$ 
We call each such map a \emph{curve diagram},
or more specifically, a curve diagram on $N$.
The reason for this terminology is that we can reconstruct (up to isotopy)
any map $f:\StdM\to N$ from
the restriction of $f$ to the equator of $\StdM.$ % XXX XXX XXX \cggb{citation? or obvious?}
In other words, we can uniquely specify any map
$f:\StdM\to N$ by indicating the action of
$f$ on the equator. We take full advantage of this
fact in our notation: any figure of a surface
with a ``green line'' drawn therein is actually notating a 
diffeomorphism, not a surface!

Any curve diagram will act on a standard POP decomposition
of $\StdM$ sending it to a POP decomposition of $N.$
Surprisingly, the converse of this statement also holds:
any POP decomposition of $N$ (a disc with holes)
comes from some curve diagram
acting on a standard POP decomposition.
%\cggb{This is true for any $N$ that is homeomorphic to a punctured disc?}
%\simon{Right, I should be more clear about when exactly we specialize to discs.}
The proof of this is constructive, and we call this
the refactoring theorem below. %\cggb{citation? or original work?}

%\simon{show how to glue two curve diagrams together to get another curve diagram}
The operation of 
gluing of surfaces can be extended to gluing of curve diagrams as
long as we are careful with the way we identify along the seam:
the identification map needs to respect the equator of the curve diagram. 
(Keep in mind that a curve diagram is really a map of surfaces,
and so gluing two such maps involves two separate gluing operations.)

@@

We note in passing two connections to the mathematical literature.
Such curve diagrams have been 
used in the study of braid groups \cite{Dehornoy2002}, and this
is where the name comes from, although our
curve diagrams respect the base points and so could be further
qualified as ``framed'' curve diagrams.
And, we note the similarity of curve diagrams and associated
modular functor 
to the definition of a planar algebra \cite{Jones1999},
the main difference being that planar algebras allow for 
not just two but any even number of curve intersections at each hole.
%\cggb{What's a planar algebra and is there an easier place for me to learn about it than from the Jones paper?}

@@

%\subsub{Refactoring theorem.}
\subsection{Refactoring theorem}
In this section we specialize to considering planar surfaces only.
The theorem we are building towards shows that
the $R$-moves act transitively
on curve diagrams.
By transitive we mean that
given two curve diagrams $f$ and $f'$ on $N$ we can find
a sequence of $R$-moves that transform $f$ to $f'.$
%We consider surface $N$ with $n$ boundary components
%$a_1,...,a_n.$
The key idea is to consider the (directed) image of the equator under
%the curve diagrams $f$ and $f'$.
these curve diagrams.
As mentioned previously, this image is sufficient to
define the entire diffeomorphism (up to isotopy) and so we are free
to work with just this image, or \emph{curve},
and we confuse the distinction between 
a curve diagram (a diffeomorphism)
and its curve.
%which we refer to as a \emph{curve.}
%Each such image will touch all the holes of $N.$
